---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Karen Thornton"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd()
library(tidyverse)
library(agricolae)

NTL_LTER <- read.csv("/Users/karenthornton/Documents/School/Grad School/Year 1/Semester 2/EDA/Environmental_Data_Analytics_2022/Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

NTL_LTER$sampledate <- as.Date(NTL_LTER$sampledate, format = "%m/%d/%y")
class(NTL_LTER$sampledate)

#2

mytheme <- theme_light(base_size = 14)+
  theme(legend.position = "bottom",
        legend.justification = "right",
        axis.text = element_text(color= "black"))

theme_set(mytheme)


```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There is no change in temperature by depth across all lakes in July
Ha: There is a change in temperature by depth across all lakes in July


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
library(lubridate)
JulydepthAllLakes <-
  NTL_LTER%>%
  mutate(Month = month(sampledate))%>%
  filter(Month == 7)%>%
  select(lakename:daynum, depth, temperature_C)%>%
  filter(!is.na(lakename), !is.na(year4), !is.na(daynum), !is.na(depth), !is.na(temperature_C))


#5

Julylakeplot <- ggplot(JulydepthAllLakes, 
                       aes(x = depth, y = temperature_C, color = lakename))+
  geom_point()+
  geom_smooth(method = lm, color = "blue")+
  ylim(0,35)+
  xlab("Depth in Meters")+
  ylab("Temperature in Celcius")
print(Julylakeplot)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: Temperature is greater with less depth. The distribution of points suggests that a linear trend might not be correct. It looks like it follows a logistic pattern. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
Julylake.regression <- lm(data = JulydepthAllLakes, temperature_C ~ depth)
summary(Julylake.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:About 73% of variability in temperature is explained by the changes in depth. This is because the r-squared value is 0.7387. The degrees of freedom is 9726. It is statistically significant because the p-value (2.2e-16) is less than 0.05. This means that we reject the null hypothesis and prove that there is a statistically significant change in temperature by depth in all lakes. Temperature is predicted to change 1.94621 degrees for every 1m change in depth. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
JulyLakesAIC <- lm(data= JulydepthAllLakes, 
                   temperature_C ~ depth + daynum + year4)
step(JulyLakesAIC)
# all variables are good to predict temperature 

#10
Julylakes.multregression <- lm(data = JulydepthAllLakes, 
                               temperature_C ~ depth + daynum + year4)
summary(Julylakes.multregression)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables is depth, daynum, and year (don't elimate any based on the AIC). About 74% of variance is explained by the model (r-squared value is 0.7411). This is a very slight improvement over the model using only depth as the explanatory variable.  



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
#anova model
Julylakes.anova <- aov(data = JulydepthAllLakes, temperature_C ~ lakename)
summary(Julylakes.anova)

#linear model
Julylakes.lm <- lm(data = JulydepthAllLakes, temperature_C ~ lakename)
summary(Julylakes.lm)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference between mean temperature among lakes. The p-value is 2.2e-16 which is less than 0.05. This means that we reject the null. Since we reject the null there is a significant difference in means. 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
Julylakeplot2 <- ggplot(JulydepthAllLakes, 
                       aes(x = depth, y = temperature_C, color = lakename))+
  geom_point(alpha =0.5)+
  geom_smooth( aes(group = lakename, color = lakename), method = lm, se = FALSE)+
  ylim(0,35)+
  xlab("Depth in Meters")+
  ylab("Temperature in Celcius")
print(Julylakeplot2)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Julylakes.anova)
#the lakes that have different means will have a "p adj" not equal to zero

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:Central Long Lake, East Long Lake, Tuesday Lake, West Long Lake have the same mean temperature statistically. Ward lake is statistically distinct from all other lakes. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: You could use an interaction test. 


